---
title: "Hadley Wickham: Data Science in R"
---

## On Machine Learning
- If you're not using data to make decisions you're just not going to be successful
**machine learning is not a magic wand**

## On using new technologies
- When do you decide to change to a new technology:
  - read a lot of blogs, such as hacker-news for up to date information
  - often hard to evaluate new technology on existing work until after you've invested the time to learn it


## Data Science Process
- Import Data
- Tidy
- Transform
- Model - scales but doesn't surprise
- Visualize - see surprises you might not expect, but doesn't scale very well
- Communicate the results to others

### TO USE MODERN DATA ANYALYSIS YOU HAVE TO LEARN TO PROGRAM
#### Two old tensions in R programming
- It is not just a programming langue, it is also an interactive data exploration tool. This causes some "surprises" 
- There is tension between being conservative vs Utopian. Basic R is much more conservative, things change slowly.


## Principals of tidyverse
1. share data structures: Use a consistent format
2. compose simple/uniform pieces: use the pipe
3. embrace functional programming: don't write for loops when you don't have to
4. write for humans: be nice to your future self
  
 
### Data Structures
- not everything fits into the data frame structure, but most will
- data frames fit almost all needs
- sometimes you want a table like data frame that holds more complex structures, like saving models to a table with their RMSE values.
- In these cases you can use a *TIBBLE*

What if you have a mix of structures:
- IE. training data + test data -> mode -> predictions -> rmse
- Not limited inside a data frame to not just storing numbers but also COMPLEX OBJECTS
- use tibble to put model in a list form 



## Tibble
- lazy
- work better with list columns
- can hold complex objects, like models or other data frames


## Compose simple/uniform pieces
- use the pipe!!
- pipe allows to write functions and arrange them in a way that is easy to understand
- thinking about code explicitly as a form of communication
- especially helpful so you don't have to curse past you, always think about future you
EX.

```
bop_on(
  scoop_up(
  hop_through(foo_foo, forest),
  field_most
),
head
)
)
```

is much more confusing then

```
foo_foo %>%
  hop_through(forest) %>%
  scop_up(field_mouse) %>%
  bop_on(head)
)
```


## Functional Programing
- the first time you make a recipe it's good have very explicit text
- for loops emphasize on the nouns, when we usually care about the verbs
```{r}
library(purrr)
map_dbl(mtcars, mean)
map_dbl(mtcars, median)
```

Other fun functional programming functions!
```{r, eval=F}
invoke_map()
pwalk()
```

## Write for Humans
- data analysis should be a little bit fun  
- embrace exquisite language 
- humans are pack animals (use stickers, join clubs, meet people)
- start with the book!

## Rule of thumb: You should never copy and paste code, more than twice
- write it to a function
- or use a loop of some kind
