---
title: 'Introduction to R: The Very Basics'
author: "Erin Grand"
date: "Oct 14, 2017"
output:
  html_notebook:
    theme: cerulean
    toc: yes
---

# Installation
## R
- Download latest base R exe file from [CRAN](https://cran.r-project.org/bin/windows/base/)

## Rstudio
- Go to [rstudio.com](https://www.rstudio.com/products/rstudio/download/)
- Select correct install file for your PC, and install

```{r, echo=F, message=FALSE, warning=FALSE}
library(tidyverse)
```
# Begin Coding:

## Everyone's favorite first line of code: "Hello World"
```{r}
print( "Hello world!" )
```
# Assignment 

Why does this code not work?

```{r, eval=T}
my_variable < - 10
my_variable
```
Look carefully! (This may seem like an exercise in pointlessness, but training your brain to notice even the tiniest difference will pay off when programming.)


# Comments
```{r}
# This is a comment
#### This is also a comment

############################################################
# I sometimes write comments like this to seperate my code #
############################################################
```




# Data Types in R
Data type | Example
------------- | -------------
Integer | 1
Logical | TRUE
Numeric | 1.1
String / character  | "Red"
Factor (enumerated string) | "Amber" or 2 in c("Red","Amber","Green")
Complex | i
Date | "2015-04-24"


### Single Values

#### Numbers: Double and Integers
```{r}
1.5
typeof(1.5)
as.integer(42)
typeof(as.integer(42))
```

#### Boolean values: TRUE (T) or FALSE (F) 
- For multiple conditions & is and and | is or
- Equals: ==
- Not Equals: !=
```{r}
FALSE
F
TRUE
T
5 > 43
42 <= 18 & 18 > 10
42 <= 18 | 18 > 10
```

#### Characters
designated as such by the quotes
```{r}
"My name in Erin"
"52%"
typeof("52%")
```

#### Factors
Some values you want to have levels instead of individual values. For example, survey results on a 1-5 scale. 
```{r}
factor(c("Big", "Semi-Big", "Semi-Small", "Small", "Tiny"))
```
We'll talk more about factors at a later time. For now, **TRY NOT TO USE THEM UNLESS NEEDED.** They can be super confusing and annoying.


#### Data Type Functions
Function | Use
------------- | -------------
is.[data type] | Whether a vector is of a particular type
as.[data type] | Attempts to coerce a vector to a data type
str | Structure of an object including class/data type, dimensions 
class | The class(es)/data type(s) an object belongs to
summary | Summarizes an object 
dput | Get R code that recreates an object
unlist | Simplify a list to a vector
dim | Dimensions of a data type


### Compound data types
Data type | Info | Construction example(s)
------------- | ------------- | -------------
Vector | A 1D set of values of the same data type | `c(1,"a")` , `1:3`
Matrix | A 2D set of values of the same data type | `rbind(1:5,2:6)`
Array | An nD set of values of the same data type | `array(LETTERS, c(13,2))`
Data.frame  | A 2D set of values of different data types | `data.frame(a=1:26, b=11:36)`
List | A collection of objects of various data types | `list(vector=c(1,"a")`, `df=data.frame(a=1:6))`

#### Vectors
```{r}
c(1, 2, 3, 4, 5)
seq(0, 4, .25)
1:5
```
#### This is also a vector!
```{r}
c("Hello", "world")
```

*What are the data types of these vectors? Are they the same?*
Can I combine data types in one vector?

```{r}
c(1, 4, FALSE, "Erin")
```
No! Notice how R turned all of the values into CHARACTERS even if they were numbers.


*To get a particular value out of a vector, you can access it's index.*
```{r}
x <- 5:10
x
x[2]
```




#### Lists
```{r}
list(1:5, 2:4)
```
Notice that I've made a list of vectors! A list of lists (or vectors). When each of the vectors is the same length can be easily converted to a table, ie. a data frame!


#### Data Frames and Data Tables
- Each observation is it's own row
- Each column represents a measure or characteristic
```{r}
x <- list(c(1:5), c(2:6))
x
data.frame(x)
#or as from vectors, the most common way to use data frames
data.frame(x = c(1:3), y = c(3:5))
```
Normally you'll be loading in your own data frame from a CSV or excel sheet, or forming your own from vectors or lists.


# Iris Dataset: 
*Let's look at a toy dataset to play with*
We will assign the first few rows of the Iris data set to a toy data frame to play with

```{r}
# assign the first 6 rows of the iris dataset as a toy dataframe
df <- head(iris)
df

# inspect the structure of the data frame
str(df)

# compute summary statistics for each column
summary(df)
```
You can also use mean(), max(), min()...etc to look at the statistic of a particular list.
```{r}
mean(df$Petal.Width)
```

#### Indexing Data Frames: Look at a set of particular rows or columns
```{r}
# look at the fifth row
df[5, ]
```

```{r}
# three different ways to look at the second column
# by index
df[ , 2]
df$Sepal.Width
df[2]


```

#### We can even filter the dataframe to only look at particular values 
```{r}
# find rows with Sepal.Length at least 5
df$Sepal.Length >= 5

# select all rows with Sepal.Length at least 5
df[df$Sepal.Length >= 5, ]

# nicer syntax for the same thing
# note: the second entry of subset is a logical vector
subset(df, Sepal.Length >= 5)
```

#### You can filter two conditions by using multiple arguments in the filter function.
```{r}
# filter ANDs conditions when given multiple arguments
filter(df, Sepal.Length >= 5, Petal.Length <= 1.4)
```



#### Looking back at the full dataset, let's see how many Species are there.
```{r}
# get a list of all unique species
unique(iris$Species)
```
If we just wanted to get the number of species, we could look at the length of this vector.
```{r}
length(unique(iris$Species))
```


#### count the number of rows for each species
```{r}
table(iris$Species)
```


#### you can also count the number of rows for each species that have a sepal width longer than 3
```{r}
table(iris$Species, iris$Sepal.Width > 3)
```


# Vector Operations
How do I add 10 to this vector?
```{r}
1:10
```


# Loops:
### IF, IF...ELSE... and IFELSE

#### If something is true, do this.
```{r}
i <- 42
if(i > 10){print(i)}
```
#### If something do this, otherwise do this instead:
```{r}
if(i < 10) 
{ 
    print(i)
}else 
{ 
  print("nope") 
}
```
#### You can also use a similar function to the IF in excel: IFELSE
This is very helpful to create columns in tables. I.E. if a student is or is not an ES grad.

```{r}
x <- seq(1, 10, .5)
ifelse(x == floor(x), 1, 0)
```

```{r}
df <- data.frame(x = 1:50, y= sample(1:100)[1:50])
df
df$yinx <- ifelse(df$y %in% df$x, 1, 0)
df
```


### For Statement and Apply Functions
A traditional for loop looks like:

```{r}
for(i in 1:10){
  print(i)
}
```
For loops are VERY SLOW because they do the same thing over and over again. R has a BETTER way to do this. `map_?()` is a function in the PURR package. We can discuss this more later. For now, a quick exmaple:

```{r}
map_int(1:10, print)
```

## Help
Google is your number one friend!
You can also access the correct syntax for a command in R with `help(function)` or `?function`
 
```{r, eval=F}
?mean
help(mean)
```

## Now you Try!
#### Using the *mpg* dataset, answer these questions.

1. What is the manufacturer, model, and year of the car in the 42nd row?
1. How many models of cars are in this data set? How many manufactures?
1. How many cars get over 19 mpg in the city?
1. How many miles per gallon ON AVERAGE do the cars get on the highway? In the city?
1. How many models of cars did Audi make in 1999?  




