---
title: "Data IO"
date: "March 7, 2017"
output:
  html_document:
    theme: cerulean
---

## In

Format | Functions
------------- | -------------
CSV | read.csv , `readr::read_csv` 
Excel | `readxl::read_excel`
Database | `RODBC::sqlQuery` , `DBI::dbGetQuery`
SPSS / SAS / Stata | `haven::read_[prog]`


## Out

Format | Functions
------------- | -------------
CSV | `write_csv` 
Excel | `openxlsx::write.xlsx`
Database | `RODBC::sqlSave , DBI::dbWriteTable` 
SPSS / SAS / Stata | `foreign::write.foreign`

## Examples
All of the following examples use packages from the Tidyverse. I prefer `readr` and `readxl` to their basic R counterparts. 

### Read Data In:
```{r, eval=F}
read_csv() #readr
```

### Write Data out:
```{r, eval = F}
write_csv() #readr
```


### Excel Files:
This is a helper function I've written to make sure that the column names of my data frame are usable in the future code (i.e. so that column names don't have spaces or punctuation.)

```{r, eval = F}
#readxl
f <- function(file, sheetname, cn){
  x <- read_excel(file, sheet = sheetname, na = "", col_names = cn) 
  valid_column_names <- make.names(names=names(x), unique=TRUE, allow_ = TRUE)
  names(x) <- valid_column_names
  return(x)
}
```



## Other file types
As well standard formats, there's a lot of connector packages out there, including a suite for [Hadoop](https://github.com/RevolutionAnalytics/RHadoop/wiki).
