---
title: "Introduction to R: The Basics"
author: "Erin"
date: '`r Sys.Date()`'
output: revealjs::revealjs_presentation
---

```{r setup, include=FALSE,	message = FALSE, warning = FALSE}
library(knitr)
knitr::opts_chunk$set(
	echo = TRUE
)
```

# Welcome and Introduction
<!-- CRAN: The Comprehensive R Archive Network -->


## Before we jump into it, make sure that you have Rstudio, R and the package "Tidyverse" installed.

All, set? Great let's begin!

*** 

### Which language do people use for what?
```{r, echo=F}
kable(data.frame("What do you want to do?" = c(
"Building a website",
"Web application",
"Phone Application",
"Numerical Computation",
"Stats",
"Visualization",
"Typesetting",
"Big Data"),
Language = c(
"HTML, CSS",
"Java Script",
"Swift, Objective C, Java",
"IDL, Python, Matlab",
"R, Strata",
"R, JavaScript (D3)",
"LaTeX, Mardown",
"SQL, Scala, Python")
), align = "lc")
```


# why R?

***
### What problems have you seen or come in across in the way you currently do data analyses and manipulation that you wish you could change?

Discuss with team.


## Why R?
**"Struggling through programming helps you learn" - Chester Ismay**

- **Free:** R and Rstudio are free and open sourced!
- **Reproducible:** Analyses done using R are easy to run again with updated data.
- **Collaboration:** Results are easy to share as a PDF.
- **Manipulation:** Data manipulation is much easier due to cool packages.
- **Understandable:** You can follow how new variables are created in someone else's code and to remember what past-you did.

## Why R?

- **Spot Checking:** You can also QC yourself along the way, instead of all at the end. PLUS code is often easier to QC because it'll yell at you when something goes wrong.
- **Visualizations:** Play with visualizations before throwing into a dashboard to see if it looks interesting.
- **Help:** Finding answers to your programming questions is a quick google away.
- **Ease:** You can create multiple data tables (files) in R  to be read into -tableau OR you can join them together in R and only have tableau read in one file.
- **Community:** Actively developed and a very active user community



# R History

## What is R?  

R is calculator
```
2 + 2
```

## NO!! 

R is a programming languange.  Specifically it's a programming language built for statistics.  

And that's what it's best at.

---  

- R is a dialect of the S languange which was developed by John Chambers at Bell Labs in 1976 and still exists today although hasn't changed much since 1998.  
- The philosopy behind S (and R) was to allow users to begin in an interactive enviroment that didn't explicitly feel like programming.  
- As their needs and skills grew they could move into more of the programming aspects.  This helps us understand some of why R is the way it is.



# Packages

----  

* Packages are simply bits of code, external to the core R code that are designed to perform a specific function.  
* The vast majority of the usefulness and functionality of R resides in packages.  
* These packages live in online repositories and can be installed on your own system to be used.  

----  

## Installing packages

* Packages need only be installed once, although you may have to reinstall when upgrading R or when you want to use a newer version of a package.

* To install from CRAN all one needs to do is:
```
install.packages("tidyverse")
```

## Using packages

<small>
Once installed all the functions in a package are available to be used.
</small>
```{r, eval=F, evcho=T}
library(tidyverse)
filter(iris, Species == "setosa")
```
* <small> `library()` loads the package into memory and allows you to use the functions within without naming the package directly everytime. </small>
* <small>  Technically what is happening here is that when attaching a package R puts those functions in your search path, the place R looks first for objects and functions. </small>
* <small> This may cause problems if packages have functions with the same name. R will choose the version for the package loaded last. </small>
* <small> Packages are attached in your current session and need to be attached every time you start a new session. </small>

# Other important key R functions 

## Generating and manipulating sequences

---- 

```{r, collapse=TRUE}
seq(1,10)

seq(1, 20, by = 2)

seq(from = 2, by = 2.5, length.out = 10)
```


```{r, collapse=TRUE}
rep(2, 3)

rep(1:3, 4)

rep(1:3, each = 2)

rep(c("A", "B", "C"), each = 6)

```


## Basic statistics

```{r, collapse=TRUE}
set.seed(3823)
x = sample(1:1000, size = 50, replace = TRUE)
```

```{r, collapse=TRUE}
max(x)

min(x)

range(x)

mean(x)

median(x)

sum(x)

sd(x)

var(x)
```

---- 

```{r, collapse=TRUE}
y = rnorm(x, 1, 0.2 * x) + x
plot(x,y)
```

---- 

### Simple linear model

```{r}
my_model = lm(y ~ x)
summary(my_model)
```

----

```{r}
plot(x, y, main = "Linear model")
abline(my_model, col = "red")
```
 
 
# Where to get help

## Built-in help

R-package authors are required to document their functions although this happens at a various levels of usefulness.  

* Simply type `?function_name` to get help on a function. 
* Look carefully what parameters the function requires and what type they are.  
* Some are required (listed first, no default) and some are optional (a default value is usually listed).  
* Most function help will also indicate what the function returns.  
* Good documentation also has more information on what the function is doing.


## Elsewhere

* Sometimes authors will provide more detailed documentation online.  
* This is more common for more recent packages where the authors may have a github repository and associated webpage.  
* Often discussion pages (Google groups, Stackoverflow) can also be a useful source of help

# Errors

## __GOOGLE IT!!!__

If I get an error I haven't seen, the first thing I will do is Google it. Usually within a few clicks I can find what went wrong.

---- 

**But sometimes an easy answer can't be found so here's a quick process to walk through:**

1.  Re-read the error and then think about it for a minute.  See if you can't get a grasp on what's really going wrong.
2.  Check your code for errors.  Spelling errors, misplaced commas, forgotten parenthesis can all cause problems
3.  Look it up - I very, very rarely get an error that someone else hasn't seen before.
4.  If you still can't find a solution then you can ask for help. I can answer brief questions or you can post questions online at Stack Overflow.  Sometimes package developers have specific discussion groups on Google groups or Github.  These can be very useful.

---- 

To get you started here are few of the more common errors you might see:

*Think about what is going wrong for each of these.*

----

```{r, error=TRUE}
my_object
```
  
----  
```{r, error=TRUE}
iris[, 6]
```

Hint:  How many columns does the `iris` data frame have?

---- 
```{r, error=TRUE}
sample[1:10,]
```

Hint:  What does `typeof(sample)` give you? What about `sample(10)`?  Or `?sample`


---- 
```{r, error=TRUE}
pet_a_cat()
```


----
```{r, error=TRUE}
nothing = NA
if (nothing == NA) {
	print("empty")
}
```

Hint:  What does `nothing == NA` give you?  How about `is.na(nothing)`?

---- 
```{r, error=TRUE}
my_data = read.table("mydata.txt")
```

Hint: Read the error message carefully.

----
```{r, error=TRUE}
x = data.frame(y = NULL)
x$y = 1:4
```

Hint: How many rows does `x` have?


# Questions so Far
## Any Questions you have from the tutorial?
What did you find to be the most challenging?


# Scripting

## What are the important parts of a program?
- Set working directory
- Load packages
- Load data
- Functions
- Code!


## How do you write fast, good, clean code?
- Comment comment comment
- Use tabs and spaces to the best of your ability
- THINK VERY HARD ABOUT EVERY FOR LOOP YOU USE. Do you really need it? (We'll talk about this more in a bit.)
- If you're using the same chunk of code  over and over again, write it into a function.


# Sample Script...

## Sample Script...

```{r, echo=T, eval=F, tidy=TRUE, tidy.opts=list(width.cutoff=50)}
# Set working directory
setwd("S:/Data Analytics/Resource Library/R help")

# Load packages
library(readxl)
library(tidyverse)

# Define Functions
# file = file name, sheetname = sheet number or name, cn = T or F
loaddata <- function(file, sheetname, cn){
  x <- read_excel(file, sheet = sheetname, na = "", col_names = cn)
  valid_names <- make.names(names=names(x), unique=TRUE, allow_ = TRUE)
  names(x) <- valid_names
  return(x)
}

# Load Data
statetestNYS <- loaddata("Uncommon Student Level Results 15-16.xlsx", "Student Scores", T)

# Look at historical state test profiency by school for ELA and Math in schools
plotdat <- statetestNYS %>% 
  filter(Subject != "Science", 
         Region == "NY",
         !(Grade == 8 & Subject == "Math")
         ) %>%
  mutate(prof = ifelse(Standard.Achieved %in% c("Level 3", "Level 4"), 1, 0)) %>%
  group_by(Year, Uncommon.school.abbreviation, Subject, Grade) %>% 
  summarise(avgP = mean(prof))

ggplot(plotdat, aes(Year, avgP, color=Subject)) +
  geom_boxplot() +
  facet_grid(Subject~Grade)
```

# Now you try!
***

#### Using the *mpg* dataset, answer these questions.

1. What is the manufacturer, model, and year of the car in the 42nd row?
1. How many models of cars are in this data set? How many manufacturers?
1. How many cars get over 19 mpg in the city?
1. How many miles per gallon ON AVERAGE do the cars get on the highway? In the city?
1. How many models of cars did Audi make in 1999?  



***

#### Answers!
```{r, echo=F, message=FALSE, warning=FALSE}
library(tidyverse)
df <- mpg
df[42, c(1, 2, 4)]

paste(length(unique(df$model)), length(unique(df$manufacturer)), sep= ", ")

sum(mpg$cty > 19)

paste(round(mean(mpg$hwy)), round(mean(mpg$cty)) , sep= ", ")

x <- filter(mpg, manufacturer == "audi", year == 1999)
length(unique(x$model))
```

# Tips and Tricks
## Some Tips and Tricks
>- **Use tab complete:** Press "tab" after typing the first few letters to find the function, variable, or file that you need without having to type it all in.
>- **Highlight Strings:** You can highlight what you want to put quotes around at once.
>- **Short Cuts:** Hightlight what you want to run and press CNT-ENTER to run it.
