---
title: "R Introduction Part 1"
author: "Erin"
date: "October 11, 2016"
output: ioslides_presentation

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```



## Welcome and Introduction
Before we jump into it, make sure that you have Rstudio, R and the package "Tidyverse" installed.

>- All, set? Great let's begin!

*** 

### Which language do people use for what?
```{r, echo=F}
library(knitr)
kable(data.frame("What do you want to do?" = c(
"Building a website",
"Web application",
"Phone Application",
"Numerical Computation",
"Stats",
"Visualization",
"Typesetting",
"Big Data"),
Language = c(
"HTML, CSS",
"Java Script",
"Swift, Objective C, Java",
"IDL, Python, Matlab",
"R, Strata",
"R, JavaScript (D3)",
"LaTeX, Mardown",
"SQL, Scala, Python")
), align = "lc")
```

***

#### What problems have you seen or come in across in the way you currently do data analyses and manipulation that you wish you could change?

***

### Why R?
**"Struggling through programming helps you learn" - Chester Ismay**

>- **Free:** R and Rstudio are free and open sourced!
>- **Reproducible:** Analyses done using R are easy to run again with updated data.
>- **Collaboration:** Results are easy to share as a PDF.
>- **Manipulation:** Data manipulation is much easier due to cool packages.
>- **Understandable:** You can follow how new variables are created in someone else's code and to remember what past-you did.

***
### Why R?

>- **Spot Checking:** You can also QC yourself along the way, instead of all at the end. PLUS code is often easier to QC because it'll yell at you when something goes wrong.
>- **Visualizations:** Play with visualizations before throwing into a dashboard to see if it looks interesting.
>- **Help:** Finding answers to your programming questions is a quick google away.
>- **Ease:** You can create multiple data tables (files) in R  to be read into tableau OR you can join them together in R and only have tableau read in one file.


## Any Questions you have from the tutorial?
- What did you find to be the most challenging?

## What are the important parts of a program?
>- Set working directory
>- Load packages
>- Load data
>- Functions
>- Code!


## How do you write fast, good, clean code?
>- Comment comment comment
>- Use tabs and spaces to the best of your ability
>- THINK VERY HARD ABOUT EVERY FOR LOOP YOU USE. Do you really need it? (We'll talk about this more in a bit.)
>- If you're using the same chunk of code  over and over again, write it into a function.


## Sample Script...


# Now you try!
***
#### Using the *mpg* dataset, answer these questions.

1. What is the manufacturer, model, and year of the car in the 42nd row?
1. How many models of cars are in this dataset? How many manufacturues?
1. How many cars get over 19 mpg in the city?
1. How many miles per gallon ON AVERAGE do the cars get on the highway? In the city?
1. How many models of cars did audi make in 1999?  



***

#### Answers!
```{r, echo=F, message=FALSE, warning=FALSE}
library(tidyverse)
df <- mpg
df[42, c(1, 2, 4)]

paste(length(unique(df$model)), length(unique(df$manufacturer)), sep= ", ")

sum(mpg$cty > 19)

paste(round(mean(mpg$hwy)), round(mean(mpg$cty)) , sep= ", ")

x <- filter(mpg, manufacturer == "audi", year == 1999)
length(unique(x$model))
```


## Some Tips and Tricks
>- **Use tab complete:** Press "tab" after typing the first few letters to find the function, variable, or file that you need without having to type it all in.
>- **Highlight Strings:** You can highlight what you want to put quotes around at once.
>- **Short Cuts:** Hightlight what you want to run and press CNT-ENTER to run it.
