---
title: "Introduction to R: Tidyverse"
author: "Erin Grand"
date: "Oct 14, 2017"
output: ioslides_presentation
---

```{r, echo=F, include=F}
library(tidyverse)
library(readxl)
```

# Introduction to R: Data Manipulation
Focusing on packages found in tidyverse


## Packages
Download tidyverse:
Imports: dplyr, ggplot2, lubricate, readr, readxl, stringr, tibble, tidyr, magrittr
if you're using multiple packages make sure you're loading them in the right order
```{r}
library(readxl)
library(tidyverse)
```


## Learn to think in Code
How to do you currently think about a task in excel or tableau?

***


## SPLIT APPLY COMBINE
1.	Split up the original data (this can be any format including data frames, lists, arrays, matrices, vectors),
2.	Apply existing or custom functions to it, and
3.	Combine the results in the same or different format.


##  Five Most Important DPLYR Functions/Verbs
1.	Filter
2.	Select
3.	Arrange
4.	Mutate
5.	Group by, Summarize

## There are so many things we didn't cover!
>- **Reshaping:** the data frame by making one column into 2. I.e serpate ELA and Math score columns instead of one column of score and one of subject.
>- **Plots:** GGPLOT is a part of tidyverse, and will allow you make simple plots very quickly.
>- **Regressions and Models:** R is first and foremost a statsical language. We'll cover linear and logistic regressions next time.
>- and much more... 
