---
title: "Tidy Data"
author: "Erin Grand"
date: "3/15/2017"
output: revealjs::revealjs_presentation
---


```{r, message=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(collapse = FALSE)
library(dplyr)
library(tidyr)
library(fivethirtyeight)
```


### The essence of tidy data is:
  
1. Each variable forms a column
1. Each observation forms a row
1. Each type of observational unit forms a table

-----

### Why?

- standarized
- easier for you and others to analyze
- easier for the computer to analyze
- easier to go from Tidy format into other formats if needed

-----

Example: Is this messy or tidy data?

```{r}
head(iris)
```

-----

Is this tidy data?
```{r}
gather(iris, petal_measurments, value, -Species) %>% head()
```


-----
For a more detailed summary of the tidy data concept, see the `tidyr` vignette

```
browseVignettes("tidyr")
```
------

# Cleaning and restructuring data


## Step 1:  Make sure it imported correctly  

- Are your column types correct?
- Are all the columns and rows there?
- How about your headers (if you have them)?
- Did R give you any warnings?
- Look at your data!


## Step 2:  Tidy it up

- Think about how you want your data structured for analysis
- What are your observations?  What are your variables?  Do you need multiple tables to capture your data better?
- Use `tidyr::gather()` and `tidyr::spread()` to appropriately transform your data
- If you have super messy data with lots of code needed to clean it up consider saving a cleaned up copy of your data and creating a separate script that documents everything you did.

# Analysis
## Doing useful stuff with your data

What you do next will depend on your own specific analyis.  Somethings you may do include:  

- making plots
- statistical summaries (mean, min, max, etc)
- statistical tests (Are the means different? Did my experiment work?!?!)
- add in to a larger data set such as sample data for an RNA-seq experiment


## Pipes: Ceci n'est pas un pipe

Don't forget to use your pipes!

```{r}
iris %>% 
	gather(petal_measurments, value, -Species) %>% 
	group_by(Species, petal_measurments) %>% 
	summarise(mean = mean(value))
```






